(tentative) Rotational Spectral Unmixing of Exoplanets

### 2017-02-21 paper/main.tex
minor text edits

We consider a simplified Earth map as shown in the upper left panel of Figure \ref{fig:mockdata}. 
This map is based on the land classification by the International Geosphere-Biosphere Programme (IGBP)\footnote{\url{https://climatedataguide.ucar.edu/climate-data/ceres-igbp-land-classification}}. 
Although the original classification assumes 16 land surface types plus ocean, in this paper we assume 3 surface types for simplicity regarding ``Open Shrubs'', ``Urban'', ``Snow/Ice'', and ``Barren/Desert'' as ``soil'' (red in the upper left panel of Figure \ref{fig:mockdata}) and other land surface types as ``vegetation'' (green), while keeping the ``ocean'' regions.

We consider an observation of diurnal light curves in 4 photometric bands. 
The assumed albedo spectra of these surface types are shown in the upper right panel of Figure \ref{fig:mockdata}. 
Here, the data for ocean are based on \citet{Mclinden1997}, 
and the data for soil and vegetation are taken from ASTER spectral library\footnote{\url{https://speclib.jpl.nasa.gov/}. 
Specifically, we adopt  ``Brown to dark brown sand'' for ``soil'', and ``Grass'' for the ``vegetation''}. 
These 4 photometric bands actually correspond to the 0.4-0.5 $\mu $m, 0.5-0.6 $\mu $m, 0.6-0.7 $\mu $m, and 0.7-0.8 $\mu $m, respectively, but they will be simply called by the band indices unless otherwise needed. 
The scattering phase function by the surface is assumed to obey the Lambert law, i.e., the radiance is (incident flux)$\times $(albedo)/$\pi$, independent of the direction of the scattering.

The multi-band light curves are synthesized given the relative configuration of the star, planet, and the observer.  
For the sake of simplicity in explaining the geometry, we consider a planet with zero obliquity in an edge-on orbit, and change the phase angle (the planet-centric angle between the star and the observer) denoted by $\alpha $. 
However, the discussion in the following is generic and does not depend on these assumptions.

The bottom panel of Figure \ref{fig:mockdata} displays the examples of diurnal light curves in 4 photometric bands, represented in terms of {\it apparent albedo}. 
Apparent albedo is the planetary intensity normalized by that of a loss-less Lambert sphere with the same radius and at the same phase \citep{Qiu2003, Seager2010}; in this paper we will use apparent albedo unless otherwise noted. 
The apparent albedo is straightforwardly obtained from the observed planetary intensity, if and only if the planetary radius and the observational geometry are completely known: latitude/longitude of the sub-stellar/sub-observer points, and the distance between the star and the planet. 
(We will discuss the effect of this assumptions later in Section ??.)

@@

The original data, however, have 4 photometric bands (i.e., $J=4$), and $J$-the dimensional space is inconvenient for visualization. 
We thus reduce the dimensionality using principal component analysis (PCA). 
As mentioned in \citet{Cowan2011}, the number of the spectrally distinct surface types is number of PCs plus 1 in theory. 
Reasonably, we extract two dominant principal components (PCs) through PCA (other components have virtually zero contribution) from the mock light curves. 
The PCs extracted from the light curves at $\alpha = 90^{\circ }$ (olive lines in Figure \ref{fig:mockdata}) are presented in Figure \ref{fig:PCs}. 
Note that, although the PCs from other light curves are not necessarily same as the one shown in Figure \ref{fig:PCs}, the plane they make should be the same plane (under the noiseless condition). Thus, in the following, we adopt these two PCs as the axes of the PC plane.

Figure \ref{fig:trajectory} shows the trajectories of the mock light curves with varying phase angle. 
The excursion of the trajectory is larger for the light curve at a larger phase angle, i.e., at crescent phase, and shrinks as the phase angle decreases when larger surface area are averaged. 
The points in the figure indicate the input albedo spectra of the three surface types. 
Because the apparent albedo at any time of the observations is a linear combination of input surface colors (Equation \ref{eq:tilde_d_f_ast_s}) under the conditions (\ref{eq:tilde_f_range}) and (\ref{eq:tilde_f_sum}), the trajectories should always be inside the triangle defined by these points.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \begin{center}
\includegraphics[width=\hsize]{mockdata_3types_t360_lc_PCplane.pdf}
    \end{center}
    \caption{Lines: trajectory of the 4-band light curves projected on the PC plane, in the case of $\alpha = 135^{\circ }$ (thick indigo), $\alpha = 90^{\circ }$ (olive), and $\alpha = 45^{\circ }$ (gold). Points: The input albedo spectra of ocean (blue), soil (red), and vegetation (green), projected on the PC plane. The surface albedo spectra cannot be located in the ray region, because it violates the condition (\ref{eq:tilde_s_range}). }
    \label{fig:trajectory}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

When it comes to the point where we must estimate the points given the trajectory/-ies, however, {\it any} set of $\{ {\bf s}_k \}$ that enclose the data points $\{{\bf d}_i\}$ in the hyperplane can be a solution of Equation (\ref{eq:tilde_d_f_ast_s}) subjected to the conditions (\ref{eq:tilde_f_range}) and (\ref{eq:tilde_f_sum}). 
The associated matrix, $\fast _{ik}$, can always be found. 
Additional constraint comes from the condition (\ref{eq:tilde_s_range}). This limits the range where $\{ {\bf s}_k \}$ can exist, but in general this is not sufficient to result in a unique solution. 
Therefore, predicting ${\bf \fast }$ and ${\bf s}$ from ${\bf d}$ is clearly a ill-posed problem.

@@

In order to see this more quantitatively, we made a grid on the PC plane  with an interval of 0.2 and considered all combinations of three grid points. 
Then, we assumed that all of the combinations that enclose the trajectory (in the case of $\alpha = 1356{\circ }$) are equally likely to be a solution, and found the marginalized probability of having a surface type at each location of the PC plane, as shown in Figure \ref{fig:gridsearch}. 
%
As expected, we find a relatively definite peak near the point of ocean. 
On the other hand, the peak that corresponds to soil is a blur, and it is almost not constrained.

>>>>

We consider diurnal light curves of a toy model of the Earth, in 4 photometric bands. 
We use a simplified surface map as shown in the upper left panel of Figure \ref{fig:mockdata}. 
This map is based on the land classification by the International Geosphere-Biosphere Programme (IGBP)\footnote{\url{https://climatedataguide.ucar.edu/climate-data/ceres-igbp-land-classification}}. 
Although the original classification assumes 16 land surface types plus ocean, in this paper we assume 3 surface types for simplicity regarding ``Open Shrubs'', ``Urban'', ``Snow/Ice'', and ``Barren/Desert'' as ``soil'' (red in the upper left panel of Figure \ref{fig:mockdata}) and other land surface types as ``vegetation'' (green), while keeping the ``ocean'' regions.

The assumed albedo spectra of these surface types in 4 photometric bands are shown in the upper right panel of Figure \ref{fig:mockdata}. 
These 4 photometric bands actually correspond to the 0.4-0.5 $\mu $m, 0.5-0.6 $\mu $m, 0.6-0.7 $\mu $m, and 0.7-0.8 $\mu $m, respectively, but they will be simply called by the band indices unless otherwise needed. 
The albedo spectra for ocean are based on \citet{Mclinden1997}, 
and the data for soil and vegetation are taken from ASTER spectral library\footnote{\url{https://speclib.jpl.nasa.gov/}. 
Specifically, we adopt  ``Brown to dark brown sand'' for ``soil'', and ``Grass'' for the ``vegetation''}. 
The scattering phase function by the surface is assumed to obey the Lambert law, i.e., the radiance is (incident flux)$\times $(albedo)/$\pi$, independent of the direction of the scattering.

The light curves are synthesized given the relative configuration of the star, planet, and the observer.  
For the sake of simplicity in explaining the geometry, we consider a planet with zero obliquity in an edge-on orbit, and change the phase angle (the planet-centric angle between the star and the observer) denoted by $\alpha $. 
However, the discussion in the following is generic and does not depend on these assumptions.

The bottom panels of Figure \ref{fig:mockdata} displays the examples of diurnal light curves in 4 photometric bands, represented in terms of {\it apparent albedo}. 
Apparent albedo is the planetary intensity normalized by that of a loss-less Lambert sphere with the same radius and at the same phase \citep{Qiu2003, Seager2010}; in this paper we will use apparent albedo unless otherwise noted. 
The apparent albedo is straightforwardly obtained from the observed planetary intensity, if and only if the planetary radius and the observational geometry are completely known: latitude/longitude of the sub-stellar/sub-observer points, and the distance between the star and the planet. 
(We will discuss the effect of this assumptions later in Section ??.)

@@

The original data, however, have 4 photometric bands (i.e., $J=4$), and $J$-the dimensional space is inconvenient for visualization. 
We thus reduce the dimensionality using principal component analysis (PCA). 
As mentioned in \citet{Cowan2011}, the number of the spectrally distinct major surface types is number of dominant PCs plus 1 in theory. 
Reasonably, we extract two dominant principal components (PCs) through PCA (other components have virtually zero contribution) from the mock light curves prepared in Section \ref{s:mockdata}. 
The PCs extracted from the light curves at $\alpha = 90^{\circ }$ (olive lines in Figure \ref{fig:mockdata}) are presented in Figure \ref{fig:PCs}. 
Although the PCs from other light curves are not necessarily same as the one shown in Figure \ref{fig:PCs}, the plane they make should be the same plane (under the noiseless condition). Thus, in the following, we adopt these two PCs as the axes of the PC plane.

Figure \ref{fig:trajectory} shows the trajectories of the mock light curves with varying phase angle on the plane defined by PC 1 and PC 2. 
The excursion of the trajectory is larger for the light curve at a larger phase angle, i.e., at crescent phase, and shrinks as the phase angle decreases because a larger surface area are averaged. 
The points in the figure indicate the input albedo spectra of the three surface types. 
Because the apparent albedo at any time of the observations is a linear combination of input surface colors (Equation \ref{eq:tilde_d_f_ast_s}) under the conditions (\ref{eq:tilde_f_range}) and (\ref{eq:tilde_f_sum}), the trajectories should always be inside the triangle defined by these points.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
    \begin{center}
\includegraphics[width=\hsize]{mockdata_3types_t360_lc_PCplane.pdf}
    \end{center}
    \caption{Lines: trajectory of the 4-band light curves projected on the PC plane, in the case of $\alpha = 135^{\circ }$ (thick indigo), $\alpha = 90^{\circ }$ (olive), and $\alpha = 45^{\circ }$ (gold). Points: The input albedo spectra of ocean (blue circle), soil (red square), and vegetation (green triangle), projected on the PC plane. The surface albedo spectra cannot be located in the ray region, because it violates the condition (\ref{eq:tilde_s_range}). }
    \label{fig:trajectory}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

On the other hand, when it comes to the point where we must estimate the points given the trajectory/-ies, however, {\it any} set of $\{ {\bf s}_k \}$ that enclose the data points $\{{\bf d}_i\}$ in the hyperplane can be a solution of Equation (\ref{eq:tilde_d_f_ast_s}) subjected to the conditions (\ref{eq:tilde_f_range}) and (\ref{eq:tilde_f_sum}). 
Note that the associated matrix, $\fast _{ik}$, can always be found. 
Additional constraint comes from the condition (\ref{eq:tilde_s_range}). This limits the region where $\{ {\bf s}_k \}$ can exist. 
However, in general this is not sufficient to result in a unique solution. 
Therefore, predicting ${\bf \fast }$ and ${\bf s}$ from ${\bf d}$ is clearly a ill-posed problem.

@@

In order to see this more quantitatively, we made a grid on the PC plane  with an interval of 0.2 and considered all combinations of three grid points. 
Then, we assumed that all of the combinations that enclose the trajectory (in the case of $\alpha =90^{\circ }$) are equally likely to be a solution, and found the marginalized probability of having a surface type at each location of the PC plane, as shown in Figure \ref{fig:gridsearch}. 
%
As expected, we find a relatively definite peak near the point of ocean. 
On the other hand, the peak that corresponds to soil is a blur, and it is almost not constrained.
